Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-12-11 paper/paper.Rmd
Starting on revisions
WIP

---
title: "Bifurcation or state tipping: assessing transition type in a model trophic cascade"
author:
  - name: Ryan Batt
    affiliation: rb
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: espm
    footnote: Corresponding Author
address:
  - code: espm
    address: "Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720"
  - code: rb
    address: "Rensselaer Polytechnic Institute, Department of Biological Sciences, Troy, NY 12309 USA; Rutgers University, Department of Ecology, Evolution, and Natural Resources, New Brunswick, NJ 08901 USA"

abstract: |
  Ecosystems can experience sudden regime shifts due to a variety of mechanisms.  Two of the ways a system can cross a tipping point include when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that anihilates one of the stable states.  The first mechanism does not generate the phenomenon of critical slowing down (CSD), whereas the latter does generate CSD, which has been postulated as a way to detect early warning signs ahead of a sudden shift.  Yet distinguishing between the two mechanisms (s-tipping and b-tipping) is not always as straightforward as it might seem.  The distinction between "state" and "parameter" that may seem self-evident in mathematical equations depends fundamentally on ecological details in model formulation. This distinction is particularly relevant when considering high-dimensional models involving trophic webs of interacting species, which can only be reduced to a one-dimensional model of a tipping point under appropriate consideration of both the mathematics and biology involved.  Here we illustrate that process of dimension reduction and distinguishing between s- and b-tipping for a highly influential trophic cascade model used to demonstrate tipping points and test CSD predictions in silico, and later, in a natural lake ecosystem. Our analysis resolves a previously unclear issue as to the nature of the tipping point involved.

@@

Ecological systems prone to abrupt changes, such as a fishery collapsing, can be challenging to manage. These abrupt dynamics are often referred to as regime shifts, and result from nonlinearities in the underlying system dynamics. Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.  There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or a bifurcation (B-tipping).  Figure \ref{fig1} illustrates these two  mechanisms.  B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]). In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.  S-tipping can be driven by direct exogenous forcing or purely stochastic events, such as extreme events that are common in the biological time series of ecosystems [@Batt2017]. Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g. @Wissel1984; @Scheffer2009; @Scheffer2015].

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{figures/scheffer-fig.png}
\caption{S-tipping vs B-tipping. A critical transition in a system with
alternative stable states can be driven either by a perturbation to the
state, driving the system across the separatrix (top panel), or by a
bifurcation, in which a stable point is annihilated (bottom panel). 
Figure reproduced from @Scheffer2001 with permission.}
\end{figure}

B-tipping is typically considered to be the result of slow change in a parameter of [e.g. @Scheffer2009] the system which ultimately leads to the fold bifurcation (or close enough to a fold bifurcation for stochasticity to finish the job).  S-tipping involves manipulation of the state variable.  @Carpenter2011 describe driving the critical transition through the gradual addition of piscivores (largemouth Bass), which drives a transition characterized by low-piscivore density into one of high piscivore density.  If the state (piscivore density) is being manipulated directly, then is this not an example of S-tipping, and thus unsuitable to test for the presence of CSD?  Although the simplicity of ball-in-cup (potential well) based descriptions of regime shifts makes them alluring, this question is impossible to address properly without a more mathematically precise description of the system and the manipulation involved.  Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications.  Here, we provide a direct analysis of the underlying model in the piscivory alternative state system to demonstrate why these studies are indeed an example of B-tipping.

@@

Decreases in adult bass abundance result from mortality ($m_A A$) and harvest ($q$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq. 1). Planktivorous fish diffuse ($D_F$) between a refuge of size $F_o$ and a foraging arena, and the predation rate of $A$ on $F$ is controlled by $c_{FA}$ (Eq. 2). Growth in $J$ is controlled by the fecundity ($f$) of $A$, but are subject to losses via cannibalization at a rate $c_{JA}$, and interactions with $F$ which are controlled by the rate at which $J$ hide ($h$) in a refuge or become vulnerable ($v$) to predation at a rate $c_{JF}$ as they enter an implicit foraging arena (Eq. 3). Herbivorous zooplankton diffuse between surface waters and a deep water refuge (of size $H_R$) at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq. 4). Phytoplankton death is due to mortality and sinking at rate $m$ or predation by herbivores at rate $c_{HF}$, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq. 5).

@@

We performed simulations in which we initialized the 5D model at equilibrium with a low but negative harvest rate ($q=-0.001$), which can be interpreted as per capita stocking of $A$. We proceeded to linearly increase $q$ at each time step over 1020 years with a time step size of 0.01 years (102000 time steps total) until $q$ reached a value of 0.001. At each time step a small amount of additive, normally distributed noise was added to Eqs. 1-3. Values for state variables less than 0.001 were set to 0.001 to prevent ecologically impossible values resulting from our numerical approximation of continuous dynamics.  R code for these simulations is included in supplementary materials (<https://github.com/cboettig/bs-tipping).

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-5dsimulation-1.png}
\caption{Time series of the 5 state variables, with \(q\) (harvest rate
of adult bass) being increased linearly over time.}\label{fig3}
\end{figure}

These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.  At a $q$ of 0, the fish in the simulation underwent a critical transition (Figure \ref{fig4}). This transition occurred during time step 510, although an abrupt change in state was not observed until step 850. The changes in A and J were similar, but were inversely related to those in F. The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well. Overall, both bass time series decreased with increasing $q$ (abruptly when stocking converted to harvest at $q = 0$), whereas the abundance of planktivorous fish was positively correlated with $q$.

\begin{figure}
\includegraphics[width=\linewidth]{../pkgBuild/manuscript_report/figure-Amotion-ballincup-1.png}
\caption{Top row: motion of A for different values of $q$. Bottom row: the negative integral of the top row, depecting in the so-called ball-in-cup (or potential well) diagrams.} \label{fig4}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-bifurcationDiagram-1.png}
\caption{Bifurcation diagram derived from Eq 9. Solid lines are stable
equilibria, the dashed line is the unstable equilibrium or separatrix.}\label{fig5}
\end{figure}

>>>>

---
title: "Bifurcation or state tipping: assessing transition type in a model trophic cascade"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: espm
    footnote: Corresponding Author
  - name: Ryan Batt
    affiliation: rb
address:
  - code: espm
    address: "Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720"
  - code: rb
    address: "Rensselaer Polytechnic Institute, Department of Biological Sciences, Troy, NY 12309 USA; Rutgers University, Department of Ecology, Evolution, and Natural Resources, New Brunswick, NJ 08901 USA"

abstract: |
  Ecosystems can experience sudden regime shifts due to a variety of mechanisms.  Two of the ways a system can cross a tipping point include when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that annihilates one of the stable states.  The first mechanism does not generate the phenomenon of critical slowing down (CSD), whereas the latter does generate CSD, which has been postulated as a way to detect early warning signs ahead of a sudden shift.  Yet distinguishing between the two mechanisms (s-tipping and b-tipping) is not always as straightforward as it might seem.  The distinction between "state" and "parameter" that may seem self-evident in mathematical equations depends fundamentally on ecological details in model formulation. This distinction is particularly relevant when considering high-dimensional models involving trophic webs of interacting species, which can only be reduced to a one-dimensional model of a tipping point under appropriate consideration of both the mathematics and biology involved.  Here we illustrate that process of dimension reduction and distinguishing between s- and b-tipping for a highly influential trophic cascade model used to demonstrate tipping points and test CSD predictions in silico, and later, in a natural lake ecosystem. Our analysis resolves a previously unclear issue as to the nature of the tipping point involved.

@@

Ecological systems prone to abrupt changes, such as a fishery collapsing, can be challenging to manage. These abrupt dynamics are often referred to as regime shifts, and result from nonlinearities in the underlying system dynamics. Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.  There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001; @Ashwin2012]: state tipping (S-tipping) or a bifurcation (B-tipping).  Figure \ref{fig1} illustrates these two  mechanisms.  B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]). In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.  S-tipping can be driven by direct exogenous forcing or purely stochastic events, such as extreme events that are common in the biological time series of ecosystems [@Batt2017]. While real-world transitions will often be a combination of both processes, one mechanism will typically have a dominant role. For instance, near enough to the bifurcation, intrinsic noise will frequently drive the system across the tipping point before the actual bifurcation event is reached.  Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g. @Wissel1984; @Scheffer2009; @Scheffer2015].

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{figures/scheffer-fig.png}
\caption{S-tipping vs B-tipping. Both panels show the bifurcation diagram for a saddle-node or "fold" bifurcation. These plot the equilibrium points of a system on the vertical axis (e.g. adult bass density in our example) as a function of environmental conditions (parameters of the system).  Solid lines indicate stable points, dotted lines indicate unstable points.  The top panel illustrates Bifurcation Tipping: a system which begins in the desirable, "high" ecosystem state (top branch) experiences gradual degradation as the underlying environmental conditions change (increasing values along the $x$). At the F2 position, this stable node collides with the unstable saddle point (dotted line) and is annihilated, leaving the system to rapidly transition to the remaining undesirable stable state (Forward shift).  To restore the system to the desirable state, it is not sufficient to merely restore the environmental conditions observed at F2.  Instead, environmental conditions must improved all the way to F1, where another bifurcation occurs that creates a rapid backwards shift to the desirable state. This effect is known as hysteresis.  The bottom panel introduces state tipping: the transition from the desirable state to the undesirable state is triggered without changing the environmental conditions.  Instead, a large perturbation is introduced directly to the ecosystem state, sufficient to push the system from one basin-of-attraction to another.  Such transitions are characterized by large, sudden shocks to the state variable, rather than slow, gradual change to the environmental conditions.  These transitions do not exhibit hysteresis -- a sufficiently large perturbation could restore the system.  Figure reproduced from Scheffer et al (2001) with permission.}\label{fig2}
\end{figure}

B-tipping is typically considered to be the result of slow change in a parameter of [e.g. @Scheffer2009] the system which ultimately leads to the fold bifurcation (or close enough to a fold bifurcation for stochasticity to finish the job).  S-tipping involves manipulation of the state variable.  @Carpenter2011 describe driving the critical transition through the gradual addition of piscivores (largemouth Bass), which drives a transition characterized by low-piscivore density into one of high piscivore density.  If the state (piscivore density) is being manipulated directly, then is this not an example of S-tipping, and thus unsuitable to test for the presence of CSD?  Although the simplicity of ball-in-cup (potential well) based descriptions of regime shifts makes them alluring, this question is impossible to address properly without a more mathematically precise description of the system and the manipulation involved.  Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications of this classic lake ecosystem experiment [@Carpenter2008; @Carpenter2011].  Here, we provide a direct analysis of the underlying model in the piscivory alternative state system to demonstrate why these studies are indeed an example of B-tipping.

@@

Decreases in adult bass abundance result from mortality ($m_A A$) and harvest effort ($q$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq. 1). Stocking the lake corresponds to $q < 0$.  This continuous-time formulation for harvest or stocking effort can be justified when carried out throughout the year at a fixed rate, as in the experiments considered here, rather than in large shocks.  Planktivorous fish diffuse ($D_F$) between a refuge of size $F_o$ and a foraging arena, and the predation rate of $A$ on $F$ is controlled by $c_{FA}$ (Eq. 2). Growth in $J$ is controlled by the fecundity ($f$) of $A$, but are subject to losses via cannibalization at a rate $c_{JA}$, and interactions with $F$ which are controlled by the rate at which $J$ hide ($h$) in a refuge or become vulnerable ($v$) to predation at a rate $c_{JF}$ as they enter an implicit foraging arena (Eq. 3). Herbivorous zooplankton diffuse between surface waters and a deep water refuge (of size $H_R$) at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq. 4). Phytoplankton death is due to mortality and sinking at rate $m$ or predation by herbivores at rate $c_{HF}$, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq. 5).

@@

We performed simulations in which we initialized the 5D model at equilibrium with a low but negative harvest rate ($q=-0.001$), which can be interpreted as per capita stocking of $A$. We proceeded to linearly increase $q$ at each time step over 1020 years with a time step size of 0.01 years (102000 time steps total) until $q$ reached a value of 0.001. At each time step a small amount of additive, normally distributed noise was added to Eqs. 1-3, truncated to non-zero values. R code for these simulations is included in supplementary materials (<https://github.com/cboettig/bs-tipping>).

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{../analysis/manuscript_report/figure-5dsimulation-1.png}
\caption{Time series of the 5 state variables, with \(q\) (harvest rate
of adult bass) being increased linearly over time.}\label{fig3}
\end{figure}

These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.  At a $q$ of 0, the fish in the simulation underwent a critical transition (Figure \ref{fig3}). This transition occurred during time step 510, although an abrupt change in state was not observed until step 850. The changes in A and J were similar, but were inversely related to those in F. The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well. Overall, both bass time series decreased with increasing $q$ (abruptly when stocking converted to harvest at $q = 0$), whereas the abundance of planktivorous fish was positively correlated with $q$.

\begin{figure}
\includegraphics[width=\linewidth]{../analysis/manuscript_report/figure-Amotion-ballincup-1.png}
\caption{Top row: motion of A for different values of $q$. Bottom row: the negative integral of the top row, depicting in the so-called ball-in-cup (or potential well) diagrams.  Note that the peaks and troughs (local maxima and minima) of the potential wells, bottom row, correspond to the zeros of the rate equations (top row). Troughs indicate stable nodes, peaks indicate unstable saddle points.  Note that under sufficient stocking of bass, $q=-0.5$, the zero point is an unstable equilibrium, indicated by the peak in the potential well (bottom right).} \label{fig4}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{../analysis/manuscript_report/figure-bifurcationDiagram-1.png}
\caption{Bifurcation diagram derived from Eq 9. Solid lines are stable
equilibria, the dashed line is the unstable equilibrium or separatrix.}\label{fig5}
\end{figure}
